---
title: "Application of Dan Moore's spatial hydrology functions"
author: "Stefan Gronsdahl, Robert Chlumsky"
date: "December 15, 2019"
output:
  html_document:
    number_sections: no
    theme: cerulean
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis presents the application of Dan Moore's code for data from Camp Creek, BC. This vignette has been updated to use the functions from within the CSHShydRology package. Note that in order to run the SAGA tools, you will need to have the OSGeo4W Network Installed on your system, see the [QGIS Downloads page](https://qgis.org/en/site/forusers/download.html).

We will start by downloading the sample data for Shannon Creek Falls from [Dan Moore's link]("https://ibis.geog.ubc.ca/~rdmoore/rcode/ShannonFallsMapData.zip"). The snippet below will save the zip file to a temporary file for reference within our examples.

```{r download data. echo=T}
temp <- tempfile()
download.file("https://ibis.geog.ubc.ca/~rdmoore/rcode/ShannonFallsMapData.zip", temp)
# system("tar -xzvf ShannonFallsMapData.zip")
tempdir_ShannonFalls <- tempdir()
unzip(temp, exdir=tempdir_ShannonFalls)
```

Now we can setup our workspace, and load the `CSHShydRology` library.

```{r, echo = T}
# load libraries
library(CSHShydRology)
library(raster)

# create and set wd
wd <- paste0(getwd(),"/working_spatial_demo")
ch_create_wd(wd)
setwd(wd)

# read in raster (from our temp info)
# dem <- raster("DEM.tif") 
dem <- raster::raster(paste0(tempdir_ShannonFalls,"/ShannonFallsDEM.tif"))
```

# Ch_create_wd and ch_clear_wd

Test generating and clearing working directories. In this case we will just use the temporary directory, but it may be useful to catalogue all the spatial files into one working directory for SAGA.

```{r, echo = T}
temp_wd <- paste0(getwd(),"/test")

# test create wd function
ch_create_wd(temp_wd)

# create a temporary file, list files
file.create(sprintf("%s/new_text_file.txt",temp_wd))
list.files(temp_wd)

# test remove wd function
ch_clear_wd(temp_wd)
```

# ch_saga_fillskins

Fill the sinks in the DEM file in a number of ways.
```{r, echo = T}
# test out the three functions using all three algorithms
dem_ns1 <- ch_saga_fillsinks(dem, saga_wd = wd, "planchon.darboux.2001")
dem_ns2 <- ch_saga_fillsinks(dem, wd, "wang.liu.2006")
dem_ns3 <- ch_saga_fillsinks(dem, wd, "xxl.wang.liu.2006")

# plot filled dems
plot(dem_ns1)
legend("topleft", legend = "planchon.darboux.2001")

plot(dem_ns2)
legend("topleft", legend = "wang.liu.2006")

plot(dem_ns3)
legend("topleft", legend = "xxl.wang.liu.2006")
```


# carea

Generate the contributing area grid.
```{r, echo = T}
carea <- ch_saga_carea(dem_ns3, wd)

plot(carea)
```

# ch_saga_catchment

Note: Some errors in code were found (line numbers refer to attached "ch_saga_2019Sept03_fnonly.r" script)
- line 287 change outlet_label to 'labels'
- line 242 change 'outlets' to 'outlet'
- line 259 change 'outlets to 'outlet

Generate the catchment boundaries.
```{r, echo = T}
# input basin points for outlets - there are 11 points in this file
basin_points = st_read("basin_points.shp")
basin_points = st_transform(basin_points, "+proj=lcc +lat_1=77 +lat_2=49 +lat_0=49 +lon_0=-95 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs") 

# try function with full basin points file 
catchments = ch_saga_catchment(dem = dem_ns1, saga_wd = wd, outlet = basin_points,
                               carea_flag = 1, carea = NULL, outlet_label = NULL)

# plot output
plot(catchments$geometry)
```

# ch_saga_channel

Generate the channel within the catchment.
```{r, echo = T}

# try under both situations
channel_list <- ch_saga_channels(dem_ns1, wd, carea_flag = 1, 
                                 initmethod = 2, initvalue = 5e4,
                                 divcells = 5, minlen = 50)

channel_list <- ch_saga_channels(dem_ns1, wd, carea_flag = 0, carea = NULL, 
                                 initmethod = 2, initvalue = 5e4,
                                 divcells = 5, minlen = 50)

# read in outputs
channels = st_read("working/channels.shp")
ntwrk <- raster::raster("working/ntwrk.sdat", format = "SAGA")
route <- raster::raster("working/route.sdat", format = "SAGA")

# plot output
plot(channels$geometry)
plot(ntwrk)
plot(route)


```

# ch_contours

```{r, echo = T}

contours = ch_contours(dem)
plot(contours$geometry)


```


# ch_check_catchment 

```{r, echo = T}

ch_checkcatchment(catchment = catchments, dem = dem_ns1, outlet = basin_points,
                  outlet_label = catchments$label)

```

# ch_check_channels 

```{r, echo = T}
channels <- channel_list$channels
ch_checkchannels(dem_ns1, channels, outlet = basin_points)

```
