---
title: "Application of Dan Moore's spatial hydrology functions"
author: "Stefan Gronsdahl"
date: "December 15, 2019"
output:
  html_document:
    number_sections: no
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis presents the application of Dan Moore's code for data from Camp Creek, BC.  

```{r, echo = T}
# Clear workspace
# define functions
source("ch_saga_2019Sept03_fnonly.r")
# create wd 
wd <- "C:/cshshydrology/spatial_functions/working"
# read in raster
dem <- raster("DEM.tif")
```

# Ch_create_wd and ch_clear_wd

```{r, echo = T}
# test create wd function
ch_create_wd(wd)
# test remove wd function
ch_clear_wd(wd)
# install wd again
ch_create_wd(wd)
```

# ch_saga_fillskins
```{r, echo = T}
# test out the three functions using all three algorithms
dem_ns1 <- ch_saga_fillsinks(dem, saga_wd = wd, "planchon.darboux.2001")
dem_ns2 <- ch_saga_fillsinks(dem, wd, "wang.liu.2006")
dem_ns3 <- ch_saga_fillsinks(dem, wd, "xxl.wang.liu.2006")
# test out if input data is let out
ch_saga_fillsinks(wd, "planchon.darboux.2001")
ch_saga_fillsinks(dem, "planchon.darboux.2001")
ch_saga_fillsinks(dem, wd)
# plot output
# windows(height = 10, width = 7)
# par(mfrow = c(3, 1), mar = c(2, 2, 1, 1))
plot(dem_ns1)
legend("topleft", legend = "planchon.darboux.2001")
plot(dem_ns2)
legend("topleft", legend = "wang.liu.2006")
plot(dem_ns3)
legend("topleft", legend = "xxl.wang.liu.2006")
```


# carea

```{r, echo = T}
carea <- ch_saga_carea(dem_ns3, wd)
# plot output
# windows(height = 5, width = 5)
# par(mfrow = c(1, 1), mar = c(1, 1, 1, 1))
plot(carea)
```

# ch_saga_catchment

Note: Some errors in code were found (line numbers refer to attached "ch_saga_2019Sept03_fnonly.r" script)
- line 287 change outlet_label to 'labels'
- line 242 change 'outlets' to 'outlet'
- line 259 change 'outlets to 'outlet


```{r, echo = T}
# input basin points for outlets - there are 11 points in this file
basin_points = st_read("basin_points.shp")
basin_points = st_transform(basin_points, "+proj=lcc +lat_1=77 +lat_2=49 +lat_0=49 +lon_0=-95 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs") 
# try function with full basin points file 
catchments = ch_saga_catchment(dem = dem_ns1, saga_wd = wd, outlet = basin_points,
                               carea_flag = 1, carea = NULL, outlet_label = NULL)
# plot output
plot(catchments$geometry)
```

# ch_saga_channel

```{r, echo = T}
# try under both situations
channel_list <- ch_saga_channels(dem_ns1, wd, carea_flag = 1, 
                                 initmethod = 2, initvalue = 5e4,
                                 divcells = 5, minlen = 50)
channel_list <- ch_saga_channels(dem_ns1, wd, carea_flag = 0, carea = NULL, 
                                 initmethod = 2, initvalue = 5e4,
                                 divcells = 5, minlen = 50)
# read in outputs
channels = st_read("working/channels.shp")
ntwrk <- raster::raster("working/ntwrk.sdat", format = "SAGA")
route <- raster::raster("working/route.sdat", format = "SAGA")
# plot output
plot(channels$geometry)
plot(ntwrk)
plot(route)
```

# ch_contours

```{r, echo = T}
contours = ch_contours(dem)
plot(contours$geometry)
```


# ch_check_catchment 

```{r, echo = T}
ch_checkcatchment(catchment = catchments, dem = dem_ns1, outlet = basin_points,
                  outlet_label = catchments$label)
```

# ch_check_channels 

```{r, echo = T}
channels <- channel_list$channels
ch_checkchannels(dem_ns1, channels, outlet = basin_points)
```