---
title: "Run DCWBM on Owikeno Basin"
author: "Joel Trubilowicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
This is a quick demo on how to use the DCWBM, and what the data should look like.  Data for the Owikeno lake watershed, in the Central Coast of British Columbia is supplied as a test cast for the package.  The model forcing data (Monthly average temperature and precipitation between 1981 and 2000) was extracted at 400m resolution using the [ClimateWNA](http://cfcg.forestry.ubc.ca/projects/climate-data/climatebcwna/) toolset from the UBC Centre for Forest Conservation Genetics.  The Owikeno lake watershed is a complex watershed containing forests, alpine areas, a lake, and glaciated areas.  The model has not been tested compared to observations, or calibrated to the Owikeno lake watershed, so accuracy is unknown and not guaranteed.

Load the package and the data:
```{r}
library(DCWBM)
data('owikeno')
```

Check the data structure here:
```{r}
str(owikeno)
```

For more information on the structure required, see ```?DCWBM```.

The DCWBM is set to use the model parameters that were initially used in the Moore, et al. (2012) paper, but can be adjusted if calibration is desired. The model by default outputs a 12 month vector of runoff, which simply are the mean of all input values, by month.

```{r}
runoff <- DCWBM(owikeno)
print(runoff)
```


If desired, the runoff can be output as a dataframe of coordinates and their runoff values.  This could then be manipulated or plotted in any way the user desires.

```{r, cache=T}
distrunoff <- DCWBM(owikeno, output = 'DISTRIB')
str(distrunoff)
```



