---
title: "At-site flood frequency analysis using threshold modeling"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{peaks-over-threshold}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---

## Introduction

This document presents how `CSHShydRology` can be used to perform at-site flood
frequency analysis using threshold modeling. 
This example is illustrated using the daily discharges of the Saint-John River 
at Fort Kent (NB).
First a verification is done to make sure that only complete year of data are 
included.
In this case, all years are complete since 1927.


```{r}
library(CSHShydRology)
data(flowStJohn)

## Verify the complete years of data 
y <- as.integer(format(flowStJohn$date,'%Y'))
ty <- tapply(y, y, length)
ty[which(ty< 365)]

## Keep only complet year
xd <- cbind(flowStJohn[y>=1927,], year = y[y>=1927])
```

## Estimating model parameters

The Function `FitPot` estimates the parameter of a Generalized Pareto
distribution (GPA) assuming that the data are independent exceedances.
$$ 
F(x)= \exp\left\{- \left( 1-\kappa \frac{x}{\alpha} \right)^{1/\kappa} \right\}
$$
The recommended estimation method is normally the maximum likelihood (MLE), 
but the L-moment method can be employed by passing the argument 
`method = 'lmom'`.
For MLE, confidence intervals of the parameters are derived using the delta 
method, while parametric bootstrap is needed for the L-moments method.


```{r}
## Random sample of GPA
rx <- rgpa(2000, 1, 0)

## Fitting POT using MLE
fit <- FitPot(rx)
print(fit)
```


In some occasions symmetrical confidence intervals can be unrealistic.  
In this case, one should prefer the confidence intervals provided 
by the profile likelihood method as returned by the function `coef`.


```{r}
## Show assymetrical confidence intervals
round(coef(fit, ci = TRUE),3)
```


## Declustering

In practice, daily streamflow evolve smoothly and declustering techniques are 
necessary to extract independents flood peaks above a given threshold $u$. 
Without declustering, exceedances are likely to be found in clusters.
Two functions are available to extract independent peaks: `which.cluster` and 
`which.floodPeaks`. 
The first method is a running declustering method that extracts the maximum of 
each cluster separated by at least `r` observations below the threshold. 
Such method may be more appropriate for extracting rainfall events. 
The second method follows the recommendation of the Water Resource Council of 
the United States as described in Lang et al. (1999). 
In brief, two adjacent peaks must respect the following two conditions: 
(i) they must be separated by $4+log(A)$ days, where $A$ is the drainage area 
in square kilometers; and (ii) the minimal intermediate flow must be less than 
75% of the lowest of the two peaks.
The example below illustrates the extraction of the peaks by the two methods. 
It is seen that the choice of a declustering method has an impact
on the extracted peaks.

```{r fig.height = 4,fig.width = 6}

## Define a threshold and the minimum separating time. Drainage Area =  14700.
thresh <- 500
r0 <- round(4 + log(14700)) # 14 days

## Extract peaks based on run declustering
peaks1 <- which.clusters(flow~date, xd, u = thresh, r = 5)

## Extract peaks based on WRC recommendations
peaks2 <- which.floodPeaks(flow~date, xd, u = thresh, r = r0, rlow = 0.75)

## Plot the peaks extracted for the year 1927
plot(flow~date, xd[xd$year == 1927,],type = 'l', col = 'grey')
abline(h = thresh, col = 2 ,lwd = 2)
points(flow~date, xd[peaks1,], pch = 16, col = 'blue', cex=2)
points(flow~date, xd[peaks2,], pch = 17, col = 'red')
legend('topleft', col = c('blue','red'), pch = 16:17, 
       legend = c('Cluster','WRC'))

```

The declustering techniques are integrated to the function `FitPot` and 
are performed when the argument `declust` is passed. 
The example below leads to the identification of the same peaks as in `peaks2` 
above, before fitting a GPA distribution.  

```{r}
## Fit a POT model after declustering
fit <- FitPot(flow~date, xd, u = 1000, declust = 'wrc', r = r0)
print(fit)
```

## Predicting flood quantiles

In the previous model, the shape parameter is positive, which 
indicates that the distribution has an upper bound $u + \alpha/\kappa$.
The flood quantiles $x_T$ of a $T$ year return period event is defined as the 
quantile of probability $[1-1/(\lambda T)]$ of the exceedance distribution, where $\lambda$ is the excess rate (average number of exceedances per year or PPY).
For the GPA distribution the flood quantiles are given by
$$
x_T = u + \frac{\alpha}{\kappa}\left[ 1 - \left(\lambda T\right)^{-\kappa}
\right],
$$
and a natural estimator of $\lambda$ is the fraction of flood events over the number
of years of observations.

The flood quantiles are evaluated by the function `predict`, where
the standard deviation and confidence intervals are included by passing 
the argument `ci`.
In the example below, the profile likelihood method is used to evaluate the 
confidence intervals of the flood quantiles of 10 and 100 return years.
Other available methods are the delta method (`delta`) and 
parametric bootstraps (`boot`).


```{r}
## Predict flood quantile of return period 10 and 100 years
predict(fit, rt = c(10,100), se = TRUE, ci = 'profile')
```

## Selecting the threshold

An import aspect of threshold modeling is the selection of the threshold. 
This decision is normally based on the notion of stability, which entails 
that if $u$ is a well-chosen threshold then all the exceedances for all
thresholds $u^\ast \geq u$ follow a GPA with the same shape parameter.
The objective is therefore to find the lowest threshold that ensures such 
stability. 
One visual assessment of the threshold stability is provided by the mean 
residual life plot (MRL).
In this graph, the mean of the exceedances is plotted against a list of
candidate thresholds.
When stability is reached, the MRL plot should exhibit a linear trend.
In the figure below, the MRL plot indicates that thresholds greater than 
1000 are reasonable choices.

```{r, fig.height=4, fig.width=6}
## List of candidate thresholds
ulst <- seq(500,1500, 25)

## Mean residual life plot
PlotMrl(flow~date, xd, u = ulst, declust = 'wrc', r = r0)
```

However, due to sampling error, MRL plots does not always lead to clear cut
decision.
Another approach for selecting a threshold is by looking 
directly at the evolution of the shape parameter with respect to the thresholds.
The function `SearchThresh` return key statistics of the POT model for several 
thresholds.
The graphics below shows that above `u = 1000` the shape parameter is stable and
leads to coherent estimated flood quantiles.

```{r, fig.height = 4, fig.width = 6}
candidates <- SearchThresh(flow~date, xd, u = ulst, declust = 'wrc', r = r0,
                           verbose = FALSE)

PlotThresh(candidates,  type = c('kappa'))
```

```{r, fig.height = 4, fig.width = 6}
PlotThresh(candidates,  type = c('q2','q5','q10','q20','q50'))
```

The process of selecting a good threshold can be automatized using a goodness of
fit test like the Anderson-Darling (AD) test.
The function `SearchThresh` also extracts the p-value of that test for every 
candidate threshold. 
The figure below shows that before a threshold of approximately 900, the 
p-values of the AD test are low and does not support the hypothesis of a GPA. 
After this changing point, the p-values increased until reaching 0.5 for a 
threshold of about 950. 
Here, the value 0.5 is the upper bound imposed by the table of Choulaki and 
Stephen (2001), which is used for interpolating the p-values. 


```{r, fig.height = 4, fig.width = 6}
PlotThresh(candidates,  type = c('ad'))
```

For the common significance level like 0.05, it was found that such procedure 
often selects too low of a threshold and so, Solari et al. (2017) suggested 
to use instead the maximum p-value. 
Alternatively, Durocher et al. (2018) suggested of using a critical values 
between 0.10 and 0.25 to avoid higher probability of falsely acceptating the 
GPA (False positive).
According to this latter criteria, the lowest stable threshold among 
the candidates is found at `u = 925`, while the maximum is `u = 950`.
These values can be extracted using the function `FindThresh` with argument
`method = 'max'` or `method = 'sgn'`.
The argument `ppy` can be used to remove candidate
thresholds that have excess rate outsite a desired interval.

```{r}
## Threshold with maximum AD p-value
cvars <- c('u','ppy','ad','q50')
FindThresh(candidates, method = 'max')[,cvars]

## First threshold with AD p-value > 0.25
FindThresh(candidates, method = 'sgn', tol.sgn = 0.25, 
           ppy = c(1,3))[,cvars]

```

Another method to automatically select the threshold is to imposed a given 
excess rate. 
By passing the argument `type = 'ppy'` and `tol.ppy = 2.2`, the candidate threshold 
with the excess rate the closest to 2.2 is returned.

```{r}
## Threshold with approximately 2.2 PPY
FindThresh(candidates, method = 'ppy', tol.ppy = 2.2)[,cvars]
```

In some situation a critical p-value such as 0.25 may not be reached or the 
threshold associated to a given excess rate may lead to samples that aren't well approximated by a GPA.
Mixing criteria for automatically selecting a threshold may lead to 
a more robust procedure (Durocher et al., 2018). 
The approach `sgn` has the advantage of generally leading to
more precise estimations as it includes more peaks and ensure that GPA is a
reasonable model.  
The arguments `sgn-max` or `sgn-ppy` ask the function `FindThresh` to search 
for the `sgn` threshold and if it cannot be found, will return the alternative 
candidates.


```{r}
candidates.mod <- candidates
candidates.mod$ad <- pmin(0.2, candidates.mod$ad)

## First threshold with AD p-value > 0.25
FindThresh(candidates.mod, method = 'sgn-ppy', 
           tol.sgn = 0.25, tol.ppy = 2.2)[,cvars]

FindThresh(candidates.mod, method = 'sgn-ppy', 
           tol.sgn = 0.15, tol.ppy = 2.2)[,cvars]
```


With the approach `sgn`, the function `FindThresh` can also reject
thresholds where the relative discrepency between the flood quantile of the 
selected threshold and a reference threshold is to large.
This serve as an additional verification that flood quantiles tend remain
stable after the selected threshold.
The reference threshold is defined as the average of the flood quantiles 
associated with the 5 lowest thresholds among the candidates.
In the example below, the argument `ppy = c(1,3)` is used to impose a lower 
bound of 1 PPY which lead to a reference of about 1 PPY.
With the St-John River, the flood quantile quickly stabilize.
After a threshold of `u = 825`, the relative discrepencies between the selected
and reference threshold is less than 1%.
This behavior results from the presence of a light tails.


```{r}
## Verify the discrepencies with reference threshold ~ 1PPY
FindThresh(candidates, method = 'sgn', ppy = c(1,3), tol.sgn = 0.0, 
           qua = 'q50', tol.qua = .01)[,cvars]

FindThresh(candidates, method = 'sgn', ppy = c(1,3), tol.sgn = 0.25, 
           qua = 'q50', tol.qua = .01)[,cvars]

```


## Conclusion

In this document, it was shown how threshold modeling can be carried out  
to estimate flood quantiles for different return period.
Instructions for fitting the model (`FitPot`), evaluate the uncertainty of 
the parameters (`coef`) and derived the flood quantile were provided 
(`predict`).
Basic validation procedures based on the mean residual life plot
(`PlotMrl`) and goodness of fit tests (`GofTest`) were shown to assess the 
fitting of the model.  
Finally, automatic procedures to selected a best candidate threshold was 
discussed (`SearchThresh` and `FindThresh`).

For a general introduction to the technical aspect of threshold modeling, 
the reader is referred to Coles (2001) or Davison and Smith (1990).

## References

Choulakian, V., Stephens, M.A. (2001). Goodness-of-Fit Tests for the Generalized Pareto Distribution. Technometrics 43, 478–484. https://doi.org/10.2307/1270819

Coles, S., 2001. An introduction to statistical modeling of extreme values. Springer Verlag.

Davison, A.C., Smith, R.L. (1990). Models for Exceedances over High Thresholds. Journal of the Royal Statistical Society. Series B (Methodological) 52, 393–442.

Durocher, M., Zadeh, S.M., Burn, D.H., Ashkar, F. (2018). Comparison of automatic procedures for selecting flood peaks over threshold based on goodness-of-fit tests. Hydrological Processes 0. https://doi.org/10.1002/hyp.13223

Lang, M., Ouarda, T.B.M.J., Bobée, B. (1999). Towards operational guidelines for over-threshold modeling. Journal of Hydrology 225, 103–117. https://doi.org/10.1016/S0022-1694(99)00167-5

Solari, S., Egüen, M., Polo, M.J., Losada, M.A. (2017). Peaks Over Threshold (POT): A methodology for automatic threshold estimation using goodness of fit p-value. Water Resour. Res. 53, 2833–2849. https://doi.org/10.1002/2016WR019426

